---
title: "Abu Dhabi pilot RNA data"
author: "Katie Burnham"
date: '`r Sys.Date()`'
output: html_document
editor_options: 
  chunk_output_type: console
---

Explanatory text largely from
(http://www.bioconductor.org/help/workflows/rnaseqGene/)[http://www.bioconductor.org/help/workflows/rnaseqGene/]
and the DESeq2 manual.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(kburnhamfunctions)
# install.packages("Vennerable", repos = "http://R-Forge.R-project.org")

# Vennerable needs graph and RBGL, from bioconductor repository
# source("http://bioconductor.org/biocLite.R")
# biocLite("graph")
# biocLite("RBGL")

# Those packages also need reshape and gtools
library(reshape)
library(gtools)
library(Vennerable)
library(DESeq2)
library(dplyr)
library(ggplot2)
library(vsn)
library(pheatmap)
library(RColorBrewer)
library(BiocParallel)
register(SnowParam(4))
library(ReportingTools)
library(ggrepel)
library(biomaRt)
```

##  Input Data

The feature count matrix prepared using the CountMatrix.R script is read in, along
with sample information. The value in the i-th row and the j-th column of the 
matrix tells how many reads (or fragments, for paired-end RNA-seq) have been
assigned to gene i in sample j. It is important to never provide counts that were
pre-normalized for sequencing depth/library size, as DESeq2's statistical model 
is most powerful when applied to un-normalized counts, and is designed to account
for library size differences internally.

```{r data}
counts <- read.delim("PilotData/PilotRNA-hisat-count-data.txt")
full.s.info <- read.delim("PilotData/PilotSampleInfoExpanded.txt",
                          stringsAsFactors=F, row.names=1)
full.s.info$PatientNumber <- as.character(full.s.info$PatientNumber)

# check order of samples
all(colnames(counts) == rownames(full.s.info))
```

The object class used by the DESeq2 package to store the read counts and the 
intermediate estimated quantities during statistical analysis is the DESeqDataSet. 
This has range-base functionality, allowing for easier integration with e.g.
ChIP-seq peaks. This object must have an associated design formula expressing the
variables used in modelling. This is used to estimate the dispersions and log2 FC, and can be changed later if all the DE steps are subsequently repeated.

NB - under default settings, the final term in the model is the variable of 
interest and control level is first.

These data are therefore converted into a DESeq2 dataset with the sample type and
disease status included in the model to begin with:

```{r convert-to-DESeq2DataSet}
dds <- DESeqDataSetFromMatrix(countData=counts,
                              colData=full.s.info,
                              design= ~ SampleType + Disease)
dds$Disease <- relevel(dds$Disease, ref = "Control")
dds

# to add feature data
featureData <- data.frame(gene=rownames(counts))
mcols(dds) <- DataFrame(mcols(dds), featureData)
mcols(dds)
```

### Pre-filtering

While it is not necessary to pre-filter low count genes before running the DESeq2 
functions, there are two reasons which make pre-filtering useful: by removing rows 
in which there are very few reads, we reduce the memory size of the dds data 
object, and we increase the speed of the transformation and testing functions 
within DESeq2. Here we perform a minimal pre-filtering to keep only rows that have 
at least 10 reads total. Note that more strict filtering to increase power is 
automatically applied via independent filtering on the mean of normalized counts 
within the results function.

```{r pre-filtering}
keep <- rowSums(counts(dds)) >= 10
dds <- dds[keep,]
dds
```

### Exploratory analysis and visualisation

Many common statistical methods for exploratory analysis of multidimensional data,
for example clustering and principal components analysis (PCA), work best for data
that generally has the same range of variance at different ranges of the mean
values. When the expected amount of variance is approximately the same across
different mean values, the data is said to be homoskedastic. For RNA-seq counts,
however, the expected variance grows with the mean. For example, if one performs
PCA directly on a matrix of counts or normalized counts (e.g. correcting for
differences in sequencing depth), the resulting plot typically depends mostly on
the genes with highest counts because they show the largest absolute differences
between samples. A simple and often used strategy to avoid this is to take the
logarithm of the normalized count values plus a pseudocount of 1; however,
depending on the choice of pseudocount, now the genes with the very lowest counts
will contribute a great deal of noise to the resulting plot, because taking the
logarithm of small counts actually inflates their variance. 

As a solution, DESeq2 offers two transformations for count data that stabilize the
variance across the mean: the regularized-logarithm transformation or rlog (Love,
Huber, and Anders 2014), and the variance stabilizing transformation (VST) for
negative binomial data with a dispersion-mean trend (Anders and Huber 2010),
implemented in the vst function. For genes with high counts, the rlog and VST will
give similar results to the ordinary log2 transformation of normalized counts. For
genes with lower counts, however, the values are shrunken towards the genes’
averages across all samples. The rlog-transformed or VST data then becomes
approximately homoskedastic, and can be used directly for computing distances
between samples, making PCA plots, or as input to downstream methods which perform
best with homoskedastic data.

Which transformation to choose? The rlog tends to work well on small datasets (n <
30), sometimes outperforming the VST when there is a large range of sequencing
depth across samples (an order of magnitude difference). The VST is much faster to
compute and is less sensitive to high count outliers than the rlog. We therefore
recommend the VST for large datasets (hundreds of samples).

On this advice, I have used the VST (this also includes correction for sequencing
depth. The "blind"=TRUE argument means that the experimental design is not used to estimate global variability in the counts i.e. more appropriate for QC purposes. 

```{r vst}
vsd <- vst(dds, blind=TRUE)

# Compare before and after
meanSdPlot(assay(dds))
meanSdPlot(assay(vsd))
```

```{r pca}
pcaData <- plotPCA(vsd, intgroup = c("Disease", "SampleType"), returnData = TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(x = PC1, y = PC2, color = SampleType, shape = Disease)) +
  geom_point(size =3) +
  xlab(paste0("PC1: ", percentVar[1], "% variance")) +
  ylab(paste0("PC2: ", percentVar[2], "% variance")) +
  coord_fixed() +
  theme_bw()
```

```{r corr-matrix}
sampleDists <- dist(t(assay(vsd)))
sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$SampleType, vsd$Disease, sep = " - " )
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
```


## Differential expression analysis

There are many possible comparisons to perform here, and the experimental design 
is complex (different cell types from the same patients and controls and multiple
time points). Contrasts of interest discussed are:

-	Overlap LPS response with healthy volunteers
-	Modifier effect of MET
-	T2D vs non
-	2nd visit (post MET treatment)

### LPS response

This is comparing the CD14 cells treated with LPS to those not treated, taking
account of patient (and excluding the 2nd visit samples).

```{r lps-response}
lps.dds <- dds[, dds@colData@listData$SampleType %in% c("CD14", "CD14_LPS")]
lps.dds <- lps.dds[, lps.dds@colData@listData$Repeat.Sample == FALSE]
# 38 samples remaining (19 individuals, 2 samples each)
lps.dds$SampleType <- droplevels(lps.dds$SampleType)
lps.dds$PatientNumber <- as.factor(lps.dds$PatientNumber)
design(lps.dds) <- ~ PatientNumber + SampleType

lps.dds <- DESeq(lps.dds, parallel=TRUE)
plotDispEsts(lps.dds, main="Dispersion plot")

lps.res <- results(lps.dds, parallel=TRUE, independentFiltering=FALSE, tidy=TRUE)

lps.res.LFC <- lfcShrink(lps.dds, coef=20, parallel=TRUE)
lps.res.LFC
summary(lps.res.LFC)

lps.res$log2FoldChange <- lps.res.LFC$log2FoldChange
lps.res$lfcSE <- lps.res.LFC$lfcSE
lpsresOrdered <- lps.res[order(lps.res$padj), ]

lpsresOrdered <- lpsresOrdered[(complete.cases(lpsresOrdered)), ]
write.table(lpsresOrdered, file="PilotData_LPSResponse_All_Patient.txt", sep="\t")

sum(lpsresOrdered$padj < 0.05, na.rm=TRUE)
# 15566

boxplot(log10(assays(lps.dds)[["cooks"]]), range=0, las=2)

res05 <- results(lps.dds, alpha=0.05)
plotMA(lps.res.LFC, ylim=c(-9, 11))

d <- plotCounts(lps.dds, gene=which.min(lps.res$padj), 
                intgroup=c("SampleType", "Disease", "PatientNumber"),
                returnData=TRUE)
ggplot(d, aes(x=SampleType, y=count, color=Disease, group=PatientNumber)) + 
  geom_point(size=3) + 
  scale_y_log10() +
  theme_bw() +
  geom_line()
```

```{r lps-report}
# des2Report <- HTMLReport(shortName = 'LPSResponse_analysis_with_DESeq2_pilot',
#                      title = 'Differential expression between CD14 and CD14_LPS',
#                          reportDirectory = "./reports")
# publish(lps.res, des2Report, pvalueCutoff=0.05, n=100,
#         DataSet=lps.dds, 
#         lfc=log(1.5, 2),
#         annotation.db = "EnsDb.Hsapiens.v86",
#         reportDir="./reports")
# finish(des2Report)
```

```{r genes}
ensembl <- useEnsembl(biomart="ensembl", dataset="hsapiens_gene_ensembl",
                      GRCh="37")
lps.genes <- getBM(attributes=c('ensembl_gene_id', 'ensembl_transcript_id',
                                'hgnc_symbol', 'chromosome_name', 'start_position',
                                'end_position'), 
                   filters = 'ensembl_gene_id', 
                   values =lpsresOrdered$row,
                   mart = ensembl)
lpsresOrdered$Gene <- lps.genes$hgnc_symbol[match(lpsresOrdered$row, 
                                                lps.genes$ensembl_gene_id)]
```

```{r volcano-plots}
lpsresOrdered$sig <- ifelse((lpsresOrdered$padj < 0.05 & 
                            abs(lpsresOrdered$log2FoldChange) >= log(1.5, 2)), 
                          "Sig DE", "Not Sig")
ggplot(data.frame(lpsresOrdered), aes(log2FoldChange, -log10(padj))) +
  geom_point(aes(col=sig)) +
  scale_color_manual(values=c("black", "red")) +
  theme_bw() +
  geom_text_repel(data=subset(data.frame(lpsresOrdered), padj<5e-250),
                  aes(label=Gene))
```

```{r lps2-correlation}
correlation.plot <- function (limma.1, limma.2, name.1 = "A", name.2 = "B", 
          main.title = "Correlation Plot", 
          key = TRUE, col.1 = "darkgreen", col.2 = "gold", col.both = "blue", 
  labels = FALSE) {
  
  limma.1 <- limma.1[complete.cases(limma.1), ]
  limma.2 <- limma.2[complete.cases(limma.2), ]
  
  limma.2 <- limma.2[which(limma.2$Gene %in% limma.1$Gene), 
    ]
  limma.1 <- limma.1[which(limma.1$Gene %in% limma.2$Gene), 
    ]
  limma.2 <- limma.2[match(limma.1$Gene, limma.2$Gene), 
    ]
  
  plot.range <- max(abs(c(limma.1$log2FoldChange, limma.2$log2FoldChange)))
  corr.plot.colours <- rep("gray", nrow(limma.1))
  for (i in 1:nrow(limma.1)) {
    if (abs(limma.1$log2FoldChange[i]) > log2(1.5) & 
        limma.1$padj[i] < 0.05 & 
        abs(limma.2$log2FoldChange[i]) > log2(1.5) & 
        limma.2$pad[i] < 0.05) {
      corr.plot.colours[i] <- col.both
    }
    else if ((abs(limma.1$log2FoldChange[i]) > log2(1.5) & 
              limma.1$padj[i] < 0.05) & 
             (abs(limma.2$log2FoldChange[i]) < log2(1.5) | limma.2$padj[i] > 0.05))
      {
      corr.plot.colours[i] <- col.1
    }
    else if ((abs(limma.1$log2FoldChange[i]) < log2(1.5) | 
              limma.1$padj[i] > 0.05) & 
             (abs(limma.2$log2FoldChange[i]) > log2(1.5) & limma.2$padj[i] < 0.05))
      {
      corr.plot.colours[i] <- col.2
    }
    if (limma.1$log2FoldChange[i] * limma.2$log2FoldChange[i] < 0) {
      if (abs(limma.1$log2FoldChange[i]) > log2(1.5) & limma.1$padj[i] < 0.05 &
          abs(limma.2$log2FoldChange[i]) > log2(1.5) & limma.2$padj[i] < 0.05) 
        {
        corr.plot.colours[i] <- "red"
      }
    }
  }
  
  plot(limma.1$log2FoldChange, limma.2$log2FoldChange, 
       xlab = paste("Log2 Fold Change", name.1), 
       ylab = paste("Log2 Fold Change", name.2), 
       main = main.title, 
       xlim = c(-plot.range, plot.range), 
       ylim = c(-plot.range, plot.range), 
       col = corr.plot.colours, pch = 20)
  
  text(x = (plot.range - 1), y = -(plot.range - 1), paste("Pearson's r =", 
    format(cor(limma.1$log2FoldChange, limma.2$log2FoldChange), digits = 3)))
  corr.test <- cor.test(limma.1$log2FoldChange, limma.2$log2FoldChange)
  if (as.numeric(corr.test$p.value) > 0) {
    text(x = (plot.range - 1), y = -(plot.range - 0.7), 
      paste("p =", format(corr.test$p.value, digits = 3)))
  }
  else {
    text(x = (plot.range - 1), y = -(plot.range - 0.7), 
      "P < 2.2 x 10-16")
  }
  if (key == TRUE) {
    legend("topleft", c("DE in Both", paste("DE in", name.1), 
      paste("DE in", name.2), "DE in Opposite Directions", 
      "Not DE"), bty = "n", col = c(col.both, col.1, col.2, 
      "red", "gray"), pch = rep(20, 4))
  }
  abline(h = 0)
  abline(v = 0)
  if (labels == TRUE) {
    text(limma.2$Gene, x = limma.1$log2FoldChange, 
         y = limma.2$log2FoldChange + 0.1, cex = 0.8)
  }
}

fairfax <- read.delim("PilotData/FairfaxScienceLPSDE.txt")
fairfax$Gene <- fairfax$Symbol

fairfax$log2FoldChange <- fairfax$logFC.LPS2.vs.Naive
fairfax$padj <- fairfax$adj.P.Val.LPS2.vs.Naive
fairfax.lps <- fairfax[, c("Gene", "log2FoldChange", "padj")]
fairfax.lps <- fairfax.lps[complete.cases(fairfax.lps), ]

fairfax.lps <- aggregate(fairfax.lps[, 2:3], by=list(fairfax.lps$Gene), FUN=mean)
colnames(fairfax.lps)[1] <- "Gene"
lps.de <- aggregate(lpsresOrdered[, c(3, 7)], by=list(lpsresOrdered$Gene),
                    FUN=mean)
colnames(lps.de)[1] <- "Gene"

correlation.plot(lps.de, fairfax.lps, "Pilot", "Fairfax", "LPS Response")
```

```{r lps24-correlation}
fairfax$log2FoldChange <- fairfax$logFC.LPS24.vs.Naive
fairfax$padj <- fairfax$adj.P.Val.LPS24h.vs.Naive
fairfax.lps <- fairfax[, c("Gene", "log2FoldChange", "padj")]
fairfax.lps <- fairfax.lps[complete.cases(fairfax.lps), ]

fairfax.lps <- aggregate(fairfax.lps[, 2:3], by=list(fairfax.lps$Gene), FUN=mean)
colnames(fairfax.lps)[1] <- "Gene"
lps.de <- aggregate(lpsresOrdered[, c(3, 7)], by=list(lpsresOrdered$Gene),
                    FUN=mean)
colnames(lps.de)[1] <- "Gene"

correlation.plot(lps.de, fairfax.lps, "Pilot", "Fairfax", "LPS Response")
```